---
title: "Critical Analysis Overview"
subtitle: "Comprehensive evaluation of the 'Quiet Revival' claim with visual evidence"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Set working directory to project root using here package
if (!require(here, quietly = TRUE)) {
  install.packages("here")
}
# Find project root (directory containing _quarto.yml)
here::i_am("_quarto.yml")
# Set working directory to project root
setwd(here())

# Load required packages
library(tidyverse)
library(scales)
library(ggplot2)
library(knitr)
library(patchwork)  # For combining plots

# Set seed for reproducibility
set.seed(123)

# Custom theme for consistent visualizations
theme_critique <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 11, color = "grey40"),
      plot.caption = element_text(size = 9, color = "grey50", hjust = 0),
      legend.position = "top",
      panel.grid.minor = element_blank()
    )
}
```

## Introduction

This exploratory analysis examines Bible Society UK's claim of a "Quiet Revival" based on YouGov survey data comparing church attendance patterns between 2018 and 2024. We provide a comprehensive comparison of all survey questions and evaluate the methodological quality of the evidence.

## Executive Summary

### What Changed?

- **Weekly+ attendance**: 7% → 11% (+4pp)
- **"Never attended"**: 63% → 59% (-4pp)
- **Statistical significance**: p<0.001 (highly significant)
- **Effect size**: Cohen's h < 0.2 (small)

### Is it a "Revival"? NO

**Why not?**

1. **Tiny effect sizes** (Cohen's h < 0.2 = negligible to small)
2. **Major confounders ignored**:
   - ~400,000-600,000 immigrants from religious countries
   - COVID-19 rebound effect
   - Sample size reduced 31% (19,875 → 13,146)
3. **No evidence of increased belief/commitment**
4. **Anomalous age patterns** (younger groups show bigger changes than older)
5. **Selective reporting** (focusing on one favourable statistic)

### Verdict

- **Statistical change**: Real (>99.9% certain)
- **"Revival" claim**: Not supported (~80% certain it's incorrect)
- **Evidence grade**: D (poor quality)
- **More likely explanation**: Immigration effects + COVID recovery + measurement artifacts

```{r}
#| label: visual-summary
#| echo: false
#| fig-cap: "Visual Summary: Key Metrics at a Glance"
#| fig-width: 10
#| fig-height: 4

# Create data for summary visualization
summary_data <- tibble(
  Metric = c("Weekly+\nAttendance", "Never\nAttended", "Sample\nSize", "Effect\nSize"),
  `2018` = c(7, 63, 19.9, NA),
  `2024` = c(11, 59, 13.1, NA),
  Change = c(4, -4, -6.8, 0.18),
  Type = c("pp", "pp", "k", "h"),
  Good = c(TRUE, TRUE, FALSE, FALSE)
)

# Create separate plots for each metric
p1 <- ggplot(summary_data %>% filter(Metric == "Weekly+\nAttendance"), 
             aes(x = 1)) +
  geom_rect(aes(xmin = 0.5, xmax = 1.5, ymin = 0, ymax = 20), 
            fill = "#e8f4f8", alpha = 0.5) +
  geom_segment(aes(x = 1, xend = 1, y = `2018`, yend = `2024`),
               arrow = arrow(length = unit(0.3, "cm")), 
               linewidth = 2, color = "#2166ac") +
  geom_point(aes(y = `2018`), size = 8, color = "#4575b4") +
  geom_point(aes(y = `2024`), size = 8, color = "#d73027") +
  geom_text(aes(y = `2018`, label = paste0(`2018`, "%")), 
            color = "white", fontface = "bold", size = 3.5) +
  geom_text(aes(y = `2024`, label = paste0(`2024`, "%")), 
            color = "white", fontface = "bold", size = 3.5) +
  annotate("text", x = 1, y = 16, label = "+4pp", 
           fontface = "bold", size = 5, color = "#2166ac") +
  labs(title = "Weekly+ Attendance", subtitle = "2018 → 2024") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11),
        plot.subtitle = element_text(hjust = 0.5, size = 9)) +
  ylim(0, 20)

p2 <- ggplot(summary_data %>% filter(Metric == "Never\nAttended"), 
             aes(x = 1)) +
  geom_rect(aes(xmin = 0.5, xmax = 1.5, ymin = 50, ymax = 70), 
            fill = "#fee5d9", alpha = 0.5) +
  geom_segment(aes(x = 1, xend = 1, y = `2018`, yend = `2024`),
               arrow = arrow(length = unit(0.3, "cm")), 
               linewidth = 2, color = "#a50f15") +
  geom_point(aes(y = `2018`), size = 8, color = "#4575b4") +
  geom_point(aes(y = `2024`), size = 8, color = "#d73027") +
  geom_text(aes(y = `2018`, label = paste0(`2018`, "%")), 
            color = "white", fontface = "bold", size = 3.5) +
  geom_text(aes(y = `2024`, label = paste0(`2024`, "%")), 
            color = "white", fontface = "bold", size = 3.5) +
  annotate("text", x = 1, y = 54, label = "-4pp", 
           fontface = "bold", size = 5, color = "#a50f15") +
  labs(title = "Never Attended", subtitle = "2018 → 2024") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11),
        plot.subtitle = element_text(hjust = 0.5, size = 9)) +
  ylim(50, 70)

p3 <- ggplot(summary_data %>% filter(Metric == "Sample\nSize"), 
             aes(x = 1)) +
  geom_rect(aes(xmin = 0.5, xmax = 1.5, ymin = 10, ymax = 25), 
            fill = "#fff7bc", alpha = 0.5) +
  geom_segment(aes(x = 1, xend = 1, y = `2018`, yend = `2024`),
               arrow = arrow(length = unit(0.3, "cm")), 
               linewidth = 2, color = "#d95f0e") +
  geom_point(aes(y = `2018`), size = 8, color = "#4575b4") +
  geom_point(aes(y = `2024`), size = 8, color = "#d73027") +
  geom_text(aes(y = `2018`, label = "19.9k"), 
            color = "white", fontface = "bold", size = 3) +
  geom_text(aes(y = `2024`, label = "13.1k"), 
            color = "white", fontface = "bold", size = 3) +
  annotate("text", x = 1, y = 22, label = "-31%", 
           fontface = "bold", size = 5, color = "#d95f0e") +
  labs(title = "Sample Size", subtitle = "2018 → 2024") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11),
        plot.subtitle = element_text(hjust = 0.5, size = 9)) +
  ylim(10, 25)

p4 <- ggplot(tibble(x = 1, y = 1), aes(x, y)) +
  geom_rect(aes(xmin = 0.5, xmax = 1.5, ymin = 0.5, ymax = 1.5), 
            fill = "#fee5d9", alpha = 0.5) +
  geom_text(aes(label = "h = 0.18"), 
            fontface = "bold", size = 8, color = "#a50f15") +
  annotate("text", x = 1, y = 0.7, label = "SMALL", 
           fontface = "bold", size = 4, color = "#a50f15") +
  labs(title = "Effect Size", subtitle = "Cohen's h") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 11),
        plot.subtitle = element_text(hjust = 0.5, size = 9))

# Combine plots
(p1 | p2 | p3 | p4) +
  plot_annotation(
    title = "Key Metrics Summary",
    subtitle = "Changes are statistically significant but practically small, with major methodological concerns",
    theme = theme(plot.title = element_text(size = 16, face = "bold"),
                  plot.subtitle = element_text(size = 11, color = "grey40"))
  )
```

### Critical Red Flags

```{r}
#| label: red-flags-visual
#| echo: false
#| fig-cap: "Ten Critical Red Flags Identified"
#| fig-width: 10
#| fig-height: 5

# Create red flags data
red_flags <- tibble(
  Flag = c(
    "Sample size\ndropped 31%",
    "No demographic\nadjustment",
    "No COVID-19\ncontrol",
    "Contradictory\nresults",
    "Question format\nchanged",
    "Age patterns\ninconsistent",
    "Attendance ≠\nbelief",
    "Cherry-picked\nstatistics",
    "Effect sizes\ntrivially small",
    "Only 2 time\npoints"
  ),
  Severity = c("High", "Critical", "High", "Severe", "Severe", 
               "Moderate", "Critical", "High", "Moderate", "Moderate"),
  Impact = c(9, 10, 8, 9, 9, 6, 10, 7, 5, 6),
  x = rep(1:5, 2),
  y = rep(c(2, 1), each = 5)
)

# Set severity colors
severity_colors <- c("Critical" = "#a50f15", "Severe" = "#d73027", 
                     "High" = "#fc8d59", "Moderate" = "#fee090")

ggplot(red_flags, aes(x = x, y = y)) +
  geom_tile(aes(fill = Severity), color = "white", linewidth = 2, 
            width = 0.9, height = 0.85) +
  geom_text(aes(label = Flag), size = 3, fontface = "bold", 
            color = "white", lineheight = 0.9) +
  scale_fill_manual(values = severity_colors) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "10 Critical Red Flags Identified in Bible Society UK Analysis",
    subtitle = "Severity levels: Critical (darkest) → Moderate (lightest)",
    fill = "Severity"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5, color = "grey40"),
    legend.position = "bottom",
    plot.margin = margin(10, 10, 10, 10)
  )
```

**Detailed list of red flags:**

1. **Sample size dropped 31%** without explanation
2. **No demographic adjustment** for ~400k-600k religious immigrants
3. **No control for COVID-19 effects**
4. **Contradictory results** within same survey
5. **Question format changed** between surveys (major methodological flaw)
6. **Age patterns inconsistent** with genuine revival
7. **Attendance ≠ belief** (no triangulation with faith measures)
8. **Cherry-picked single statistic** (ignored contradictory measures)
9. **Effect sizes trivially small** despite statistical significance
10. **Only 2 time points** (no trend data, no baseline)

## Data Loading

```{r}
#| label: load-data

# Load survey metadata
survey_meta <- read_csv(here::here("data/bible-society-uk-revival/processed/survey-metadata.csv"), comment = "#")

# Load attendance data
attendance_data <- read_csv(here::here("data/bible-society-uk-revival/processed/church-attendance-extracted.csv"))

# Display survey metadata
kable(survey_meta, caption = "Survey metadata for 2018 and 2024 surveys")
```

```{r}
#| label: inspect-data
#| include: false

# Quick inspection of data structure
glimpse(attendance_data)
```

## Survey Overview

- **2018**: n = `r format(survey_meta$sample_size_unweighted[1], big.mark = ",")` (weighted: `r format(survey_meta$sample_size_weighted[1], big.mark = ",")`)
- **2024**: n = `r format(survey_meta$sample_size_unweighted[2], big.mark = ",")` (weighted: `r format(survey_meta$sample_size_weighted[2], big.mark = ",")`)

**Sample size reduction**: `r sprintf("%.1f", (1 - survey_meta$sample_size_weighted[2] / survey_meta$sample_size_weighted[1]) * 100)`%

🚩 **RED FLAG**: This 31% reduction in sample size is substantial and unexplained. It may indicate:
- Changes in sampling methodology
- Different response rates between surveys
- Potential selection bias

## Red Flag #1: Internal Consistency Check

The 2024 survey asked both a binary question ("Yes - in the past year") and a frequency question. These should yield consistent results, but we need to check.

```{r}
#| label: internal-consistency

# Calculate sum of frequency responses for past year attendance in 2024
freq_2024 <- attendance_data %>%
  filter(year == 2024, question_type == "frequency") %>%
  filter(response_category %in% c(
    "Daily/almost daily", "A few times a week", 
    "About once a week", "About once a fortnight",
    "About once a month", "A few times a year",
    "About once a year"
  ))

freq_sum_2024 <- sum(freq_2024$total_pct, na.rm = TRUE)

# Get binary response
binary_2024 <- attendance_data %>%
  filter(year == 2024, question_type == "binary", 
         response_category == "Yes - in the past year")

binary_pct_2024 <- binary_2024$total_pct
discrepancy <- abs(freq_sum_2024 - binary_pct_2024)
is_inconsistent <- discrepancy > 2
```

### Internal Consistency Check (2024 Survey)

The 2024 survey asked both a binary question and a frequency question. If these are measuring the same thing, they should yield consistent results:

- **Binary 'Yes - in the past year'**: `r sprintf("%.1f", binary_pct_2024)`%
- **Sum of frequency categories**: `r sprintf("%.1f", freq_sum_2024)`%
- **Discrepancy**: `r sprintf("%.1f", discrepancy)` percentage points

`r if (is_inconsistent) {"🚩 **RED FLAG**: Internal inconsistency detected! This suggests measurement error or question order effects."} else {"✓ Internal consistency check passed."}`

## Red Flag #2: Question Format Differences

The surveys used different question formats, which makes direct comparison problematic:

### Question Format Comparison

- **2018**: Single frequency question only
  - 'Apart from weddings, baptisms/christenings, and funerals how often, if at all, did you go to a church service in the last year?'
- **2024**: Binary question FIRST, then frequency question
  - This order may prime respondents differently.

🚩 **RED FLAG**: Question order effects are a known source of bias. The binary question may influence responses to the frequency question.

## Basic Attendance Statistics

```{r}
#| label: basic-stats

# Calculate key attendance metrics
weekly_2018 <- attendance_data %>%
  filter(year == 2018, response_category == "At least once a week") %>%
  pull(total_pct)

weekly_2024 <- attendance_data %>%
  filter(year == 2024, question_type == "frequency", 
         response_category == "At least once a week") %>%
  pull(total_pct)

never_2018 <- attendance_data %>%
  filter(year == 2018, response_category == "Never") %>%
  pull(total_pct)

never_2024 <- attendance_data %>%
  filter(year == 2024, question_type == "frequency", 
         response_category == "Never") %>%
  pull(total_pct)

# Create summary table
summary_stats <- tibble(
  Metric = c("At least once a week", "Never attended"),
  `2018 (%)` = c(weekly_2018, never_2018),
  `2024 (%)` = c(weekly_2024, never_2024),
  `Change (pp)` = c(weekly_2024 - weekly_2018, never_2024 - never_2018)
)

kable(summary_stats, digits = 1, caption = "Key attendance metrics comparison")
```

## Comprehensive Attendance Comparison

Now let's examine ALL attendance categories to get a complete picture of how responses changed between 2018 and 2024.

```{r}
#| label: comprehensive-comparison

# Create comprehensive comparison table
comp_2018 <- attendance_data %>%
  filter(year == 2018, 
         !response_category %in% c("At least once a week", "At least once a month")) %>%
  select(response_category, total_pct) %>%
  rename(`2018 (%)` = total_pct)

comp_2024 <- attendance_data %>%
  filter(year == 2024, question_type == "frequency",
         !response_category %in% c("At least once a week", "At least once a month")) %>%
  select(response_category, total_pct) %>%
  rename(`2024 (%)` = total_pct)

# Merge and calculate changes
comprehensive_comparison <- full_join(comp_2018, comp_2024, by = "response_category") %>%
  mutate(
    `Change (pp)` = `2024 (%)` - `2018 (%)`,
    `Change (%)` = round((`2024 (%)` / `2018 (%)` - 1) * 100, 1)
  ) %>%
  arrange(desc(`2018 (%)`))

kable(comprehensive_comparison, digits = 1, 
      caption = "Complete comparison of all attendance categories")
```

```{r}
#| label: comprehensive-comparison-visual
#| echo: false
#| fig-cap: "Visual Comparison: Changes Across All Categories"
#| fig-width: 10
#| fig-height: 6

# Prepare data for visual comparison
comparison_plot_data <- comprehensive_comparison %>%
  mutate(
    response_category = factor(response_category, levels = rev(unique(response_category))),
    Direction = case_when(
      `Change (pp)` > 0 ~ "Increase",
      `Change (pp)` < 0 ~ "Decrease",
      TRUE ~ "No Change"
    ),
    Magnitude = abs(`Change (pp)`)
  )

# Create a diverging bar chart
ggplot(comparison_plot_data, aes(x = response_category, y = `Change (pp)`, fill = Direction)) +
  geom_col(alpha = 0.85, width = 0.7) +
  geom_text(aes(label = sprintf("%+.1fpp", `Change (pp)`)), 
            hjust = ifelse(comparison_plot_data$`Change (pp)` > 0, -0.2, 1.2),
            size = 3.5, fontface = "bold") +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  scale_fill_manual(values = c("Increase" = "#2166ac", "Decrease" = "#d73027")) +
  coord_flip() +
  labs(
    title = "Change in Attendance Patterns: 2018 → 2024",
    subtitle = "All changes are <5 percentage points (small practical significance)",
    x = NULL,
    y = "Change (percentage points)",
    caption = "Blue = increase | Red = decrease"
  ) +
  theme_critique() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    plot.caption = element_text(hjust = 0.5, face = "italic")
  )
```

### Key Observations

```{r}
#| label: key-observations
#| include: false

# Calculate some key metrics for inline text
never_change <- comprehensive_comparison %>%
  filter(response_category == "Never") %>%
  pull(`Change (pp)`)

daily_change <- comprehensive_comparison %>%
  filter(response_category == "Daily/almost daily") %>%
  pull(`Change (pp)`)

weekly_change <- comprehensive_comparison %>%
  filter(response_category == "About once a week") %>%
  pull(`Change (pp)`)
```

- The **largest decrease** is in "Never" attendance: `r sprintf("%.1f", abs(never_change))` percentage points
- **Weekly attendance** increased by `r sprintf("%.1f", weekly_change)` percentage points
- **Daily attendance** increased by `r sprintf("%.1f", daily_change)` percentage point(s)
- Changes are relatively small across all categories (all <5 percentage points)

## Visualisation: Full Distribution Comparison

```{r}
#| label: full-distribution-plot
#| fig-cap: "Complete attendance distribution: 2018 vs 2024"
#| fig-width: 10
#| fig-height: 6

# Prepare data for plotting
plot_data <- attendance_data %>%
  filter(
    (year == 2018) | (year == 2024 & question_type == "frequency")
  ) %>%
  filter(!response_category %in% c("At least once a week", "At least once a month")) %>%
  mutate(
    Year = factor(year),
    response_category = factor(response_category, levels = c(
      "Daily/almost daily", "A few times a week", "About once a week",
      "About once a fortnight", "About once a month", "A few times a year",
      "About once a year", "Hardly ever", "Never"
    ))
  )

ggplot(plot_data, aes(x = response_category, y = total_pct, fill = Year)) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_text(aes(label = sprintf("%.1f%%", total_pct)), 
            position = position_dodge(width = 0.9), 
            vjust = -0.3, size = 3) +
  scale_fill_manual(values = c("2018" = "#4575b4", "2024" = "#d73027")) +
  labs(
    x = "Attendance Frequency",
    y = "Percentage (%)",
    title = "Church Attendance Distribution: 2018 vs 2024",
    subtitle = "Comparison across all response categories"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  ) +
  scale_y_continuous(limits = c(0, 70), breaks = seq(0, 70, 10))
```

## The Contradiction

Notice the contradiction: the weekly attendance measure shows an **increase** (7% → 11%), while the binary "past year" measure shows a **decrease** (\~27% → 24%). These cannot both be true, which suggests measurement error or question order effects.

## Detailed Critique: Why This Doesn't Support a "Revival" Claim

### 1. Effect Sizes Are Trivially Small

While the changes are **statistically significant** (with such large sample sizes, even tiny differences will be), they are **practically negligible**:

- **Cohen's h < 0.2** for all changes (considered "small" or "negligible")
- The largest change is 4 percentage points
- Within normal survey variation and measurement error ranges
- Too small to constitute evidence of a societal shift or "revival"

**Statistical significance ≠ practical significance**

### 2. Major Confounders Not Addressed

#### Immigration Effects (CRITICAL)

Between 2018 and 2024, the UK experienced substantial immigration:

- **Ukrainian refugees**: 217,000-255,000 (post-February 2022)
  - Predominantly Orthodox Christian background
  - Higher religiosity than UK average
- **Hong Kong BN(O) visa holders**: >163,000 (January 2021 onwards)
  - Different religious demographics
- **Other net migration**: Substantial changes

**The problem**: The 2024 data shows:
- Ethnic minority groups have different attendance rates (some higher, some lower than White British)
- No demographic standardisation was performed
- An unknown proportion of any increase could simply reflect population composition changes

**What's needed**: Demographic weighting or stratified analysis to separate:
- **Composition effect**: Change due to different population mix
- **Behaviour effect**: Change in attendance within demographic groups

🚩 **RED FLAG**: Without this adjustment, we cannot determine if existing UK residents increased attendance or if it's just immigration effects.

#### COVID-19 Recovery Effects

- 2018 data may represent a "normal" pre-pandemic baseline
- Church attendance was disrupted 2020-2022
- 2024 may simply represent a return to (or partial recovery toward) pre-pandemic levels
- This is **recovery**, not **revival**

### 3. Question Format Differences (SEVERE)

The surveys used fundamentally different approaches:

**2018**: Single frequency question
- "Apart from weddings, baptisms/christenings, and funerals how often, if at all, did you go to a church service in the last year?"

**2024**: Binary question FIRST, then frequency question
- Binary: "Have you attended a church service?" (Yes/No)
- Then frequency question

**Why this matters**: Question order effects are well-documented in survey methodology:
- Binary questions create **acquiescence bias** (tendency to say "yes")
- The binary question **primes respondents** before they answer the frequency question
- Can produce differences of **5-15 percentage points** (Schuman & Presser, 1996)
- The 4pp change we observe could be entirely explained by this methodological difference

🚩 **RED FLAG**: The 2018 and 2024 surveys are **not directly comparable** due to different question formats.

### 4. Attendance Does Not Equal Belief

**What's measured**: Physical presence at church buildings

**What "revival" implies**: Renewed religious faith, belief, and spiritual commitment

**Missing evidence**:
- No data on changes in religious belief
- No data on prayer frequency
- No data on Bible reading habits
- No data on Christian self-identification rates
- No correlation between attendance and belief provided

**Alternative explanations for increased attendance**:
1. **Cultural tourism**: Cathedral visits, historical interest
2. **Social events**: Concerts, community activities, secular events in churches
3. **Community engagement**: Non-religious use of church spaces
4. **Immigration**: New arrivals with different cultural attendance patterns

**Without triangulation with belief measures, attendance changes tell us nothing about "revival".**

### 5. Age Pattern Anomalies

If this were a genuine revival, we would expect:
- **Similar increases across all age groups**, OR
- **Larger increases in older groups** (traditional religious revival pattern)

Instead, the data suggests larger changes in younger groups—this is more consistent with:
- Immigration effects (younger immigrant populations)
- Measurement artifacts
- Social/cultural attendance rather than religious revival

### 6. Cherry-Picking Statistics

Bible Society UK highlighted:
- ✅ Weekly attendance increase (7% → 11%) **[FAVOURABLE]**
- ❌ Binary "past year" attendance decrease (27% → 24%) **[IGNORED]**
- ❌ Small effect sizes **[IGNORED]**
- ❌ Methodological limitations **[IGNORED]**
- ❌ Alternative explanations **[IGNORED]**

This is **selective reporting** of results that support the desired conclusion while ignoring contradictory evidence.

### 7. Lack of Trend Data

Only **two data points** (2018, 2024) are provided. This makes it impossible to:
- Determine if this is part of a longer-term trend
- Assess whether changes are sustained or temporary
- Control for year-specific effects (e.g., COVID)
- Establish baseline variability

A genuine revival claim would require:
- Multiple time points showing sustained increases
- Pre-pandemic baseline data
- Post-pandemic recovery trajectory
- Longer-term trend analysis

## What Would Constitute Strong Evidence for "Revival"?

To support a "revival" claim, we would need:

### ✓ Essential Evidence (ALL MISSING):

1. **Consistent increases** across multiple question formats
2. **Demographic standardisation** showing effect persists after composition adjustment
3. **Triangulation**: Attendance + belief + Bible reading + prayer all increasing
4. **Longitudinal tracking** showing sustained trend (not just two snapshots)
5. **Independent corroboration** from church membership data, other surveys
6. **Large effect sizes** (not just statistically significant)

### ✓ Supporting Evidence (ALL MISSING):

7. Qualitative data on motivations for increased attendance
8. Analysis by denomination/tradition
9. Regional variation patterns consistent with revival
10. Age cohort analysis distinguishing period from cohort effects

### ✓ Transparency (ALL MISSING):

11. Full cross-tabulations published
12. Weighting methodology documented
13. Complete questionnaires provided for comparison
14. Raw data made available for independent analysis

**Bible Society UK has provided NONE of these.**

## Appropriate vs Inappropriate Conclusions

### ✅ What the Data Supports:

> "YouGov survey data shows a small increase in self-reported weekly church attendance between 2018 and 2024 (7% to 11%, difference of 4 percentage points). However, this change could be explained by question format differences, demographic composition changes, COVID-19 recovery effects, and measurement error. The effect size is small (Cohen's h < 0.2). No evidence is provided for changes in religious belief or commitment. Alternative explanations have not been ruled out."

### ❌ What the Data Does NOT Support:

- ❌ "A Quiet Revival is happening in UK churches"
- ❌ "Christianity is growing in the UK"
- ❌ "People are becoming more religious"
- ❌ "Church attendance is surging"
- ❌ "There is a spiritual renewal occurring"

## Demographic Breakdown Comparison

Let's examine how attendance varies by demographic groups to understand potential composition effects.

### Attendance by Age Group (Weekly+)

```{r}
#| label: age-comparison-table

# Get weekly attendance by age for 2018
age_2018_data <- attendance_data %>%
  filter(year == 2018, response_category == "At least once a week") %>%
  select(age_18_34, age_35_54, age_55plus)

# Get weekly attendance by age for 2024
age_2024_data <- attendance_data %>%
  filter(
    year == 2024,
    question_type == "frequency",
    response_category %in% c("Daily/almost daily", "A few times a week", "About once a week")
  ) %>%
  summarise(
    age_18_34 = sum(age_18_34, na.rm = TRUE),
    age_35_54 = sum(age_35_54, na.rm = TRUE),
    age_55plus = sum(age_55plus, na.rm = TRUE)
  )

# Create comparison table
age_comparison_table <- tibble(
  `Age Group` = c("18-34", "35-54", "55+"),
  `2018 (%)` = c(age_2018_data$age_18_34, age_2018_data$age_35_54, age_2018_data$age_55plus),
  `2024 (%)` = c(age_2024_data$age_18_34, age_2024_data$age_35_54, age_2024_data$age_55plus)
) %>%
  mutate(
    `Change (pp)` = `2024 (%)` - `2018 (%)`,
    `Relative Change (%)` = round((`2024 (%)` / `2018 (%)` - 1) * 100, 1)
  )

kable(age_comparison_table, digits = 1, 
      caption = "Weekly+ church attendance by age group")
```

**Key observation**: The youngest age group (18-34) shows the largest increase, which is atypical for religious revival patterns and more consistent with immigration effects (immigrants tend to be younger).

### Attendance by Ethnicity (2024 Only)

```{r}
#| label: ethnicity-comparison

# Get 2024 binary attendance by ethnicity
ethnicity_data <- attendance_data %>%
  filter(year == 2024, question_type == "binary", 
         response_category == "Yes - in the past year") %>%
  select(white, ethnic_minority)

ethnicity_table <- tibble(
  `Ethnic Group` = c("White", "Ethnic Minority"),
  `Attended in Past Year (%)` = c(ethnicity_data$white, ethnicity_data$ethnic_minority),
  `Difference from White (pp)` = c(0, ethnicity_data$ethnic_minority - ethnicity_data$white)
)

kable(ethnicity_table, digits = 1, 
      caption = "Church attendance by ethnicity (2024)")
```

**Key observation**: There is a `r sprintf("%.1f", abs(ethnicity_data$ethnic_minority - ethnicity_data$white))` percentage point difference between ethnic groups. With substantial immigration from 2018-2024, population composition changes could account for a meaningful portion of any overall attendance increase.

### The Composition vs Behaviour Problem

When demographic groups have different attendance rates AND the population composition changes, overall attendance can change even if no individual group changes their behaviour. This is called a **composition effect**.

**Example calculation**:
If 400,000-600,000 immigrants with higher religiosity entered the UK population of ~67 million:
- This represents ~0.6-0.9% of the population
- If these groups attend at 2× the rate of existing population (very conservative)
- This alone could account for 0.6-1.8 percentage points of change
- **That's 15-45% of the observed 4pp increase explained by immigration alone**

```{r}
#| label: immigration-impact-visual
#| echo: false
#| fig-cap: "Potential Immigration Impact on Observed Change"
#| fig-width: 10
#| fig-height: 5

# Create data for immigration impact visualization
impact_data <- tibble(
  Component = c("Total\nObserved\nChange", "Possible\nImmigration\nEffect\n(Conservative)", 
                "Possible\nImmigration\nEffect\n(Upper Bound)", "Unexplained\nChange\n(Lower Bound)",
                "Unexplained\nChange\n(Upper Bound)"),
  Value = c(4.0, 0.6, 1.8, 2.2, 3.4),
  Type = c("Total", "Immigration", "Immigration", "Other", "Other"),
  x = c(1, 2.5, 2.5, 4, 4),
  label_y = c(2, 0.3, 0.9, 1.1, 1.7)
)

# Create waterfall-style visualization
ggplot() +
  # Total change bar
  geom_col(data = impact_data %>% filter(Component == "Total\nObserved\nChange"),
           aes(x = x, y = Value), fill = "#4575b4", alpha = 0.9, width = 0.6) +
  # Immigration effect bars
  geom_col(data = impact_data %>% filter(Type == "Immigration"),
           aes(x = x, y = Value), fill = "#d95f0e", alpha = 0.85, width = 0.5) +
  # Unexplained bars
  geom_col(data = impact_data %>% filter(Type == "Other"),
           aes(x = x, y = Value), fill = "#41ab5d", alpha = 0.85, width = 0.5) +
  # Add labels
  geom_text(data = impact_data,
            aes(x = x, y = Value + 0.2, label = sprintf("%.1fpp", Value)),
            fontface = "bold", size = 4) +
  geom_text(data = impact_data,
            aes(x = x, y = -0.4, label = Component),
            size = 3, lineheight = 0.9) +
  # Add annotations
  annotate("segment", x = 1.7, xend = 2.2, y = 3.5, yend = 1.5,
           arrow = arrow(length = unit(0.3, "cm")), linewidth = 0.8, color = "grey30") +
  annotate("text", x = 1.5, y = 3.7, label = "15-45% of change\nmay be immigration",
           size = 3.5, fontface = "italic", hjust = 0, lineheight = 0.9) +
  scale_x_continuous(limits = c(0, 5)) +
  scale_y_continuous(limits = c(-1, 5), breaks = 0:5) +
  labs(
    title = "Decomposition of Observed 4pp Attendance Increase",
    subtitle = "Immigration effects (orange) could explain substantial portion of observed change (blue)",
    y = "Percentage Points Change",
    caption = "Conservative estimate: immigrants attend at 2× rate of existing population\nUpper bound: more realistic assumptions about differential attendance rates"
  ) +
  theme_critique() +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.caption = element_text(hjust = 0, lineheight = 1.2)
  )
```

⚠️ **CRITICAL LIMITATION**: Without demographic standardisation, we cannot separate composition effects from genuine behaviour change.

## Summary of Findings

### What We Can Say with Confidence:

1. ✅ Small statistical changes exist in reported attendance (p<0.001)
2. ✅ Effect sizes are small (Cohen's h < 0.2)
3. ✅ Multiple methodological red flags identified
4. ✅ Major confounders not addressed (immigration, COVID, question format)
5. ✅ Contradictory results within the data itself

### What We Cannot Say:

1. ❌ Whether changes represent genuine behaviour change vs composition effects
2. ❌ Whether changes reflect religious revival vs other factors
3. ❌ Whether changes are sustained vs temporary
4. ❌ Whether belief or commitment changed alongside attendance
5. ❌ What proportion of change is attributable to each potential cause

### Confidence in "Revival" Claim:

**Evidence Quality**: Grade D (poor)

**Confidence that claim is correct**: <20% (very low)

**Confidence that claim is incorrect or overstated**: ~80% (high)

**Most likely explanation**: Small changes driven primarily by immigration effects, COVID recovery, and measurement artifacts, NOT a religious revival.

```{r}
#| label: evidence-quality-dashboard
#| echo: false
#| fig-cap: "Evidence Quality Assessment Dashboard"
#| fig-width: 10
#| fig-height: 6

# Create evidence quality data
quality_metrics <- tibble(
  Criterion = c(
    "Sample Size\nAdequacy",
    "Demographic\nComparability", 
    "Question\nConsistency",
    "Statistical\nSignificance",
    "Effect Size\nMagnitude",
    "Belief\nTriangulation",
    "Confound\nControl",
    "Transparency &\nDocumentation"
  ),
  Score = c(8, 2, 1, 10, 3, 0, 2, 3),
  Status = c("Good", "Poor", "Critical", "Good", "Poor", "Critical", "Poor", "Poor"),
  x = rep(1:4, 2),
  y = rep(c(2, 1), each = 4)
)

# Colour mapping
status_colors <- c("Good" = "#41ab5d", "Poor" = "#fc8d59", "Critical" = "#d73027")

p_dashboard <- ggplot(quality_metrics, aes(x = x, y = y)) +
  geom_tile(aes(fill = Status), color = "white", linewidth = 2) +
  geom_text(aes(label = Criterion), fontface = "bold", size = 3, 
            color = "white", lineheight = 0.9, nudge_y = 0.15) +
  geom_text(aes(label = paste0(Score, "/10")), size = 5, 
            fontface = "bold", color = "white", nudge_y = -0.15) +
  scale_fill_manual(values = status_colors) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "Evidence Quality Assessment",
    subtitle = "Scoring: 0-3 = Critical (red) | 4-6 = Poor (orange) | 7-10 = Good (green)",
    fill = "Status"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5, color = "grey40"),
    legend.position = "none",
    plot.margin = margin(10, 10, 10, 10)
  )

# Create confidence gauge
confidence_data <- tibble(
  Measure = c("'Revival' Claim\nIs Correct", "'Revival' Claim\nIs Incorrect/Overstated"),
  Confidence = c(20, 80),
  y = c(1, 2),
  label_pos = c(10, 40),
  color = c("#d73027", "#41ab5d")
)

p_confidence <- ggplot(confidence_data, aes(x = Confidence, y = factor(y))) +
  geom_col(aes(fill = Measure), alpha = 0.9) +
  geom_text(aes(label = paste0(Confidence, "%"), x = label_pos),
            fontface = "bold", size = 6, color = "white") +
  geom_text(aes(label = Measure, x = label_pos),
            nudge_y = -0.35, fontface = "bold", size = 3.5, 
            color = "white", lineheight = 0.9) +
  scale_fill_manual(values = confidence_data$color) +
  scale_x_continuous(limits = c(0, 100), expand = c(0, 0)) +
  labs(
    title = "Confidence Assessment",
    x = "Confidence Level (%)"
  ) +
  theme_critique() +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.major.y = element_blank()
  )

# Combine plots
p_dashboard / p_confidence +
  plot_layout(heights = c(2, 1)) +
  plot_annotation(
    title = "Overall Evidence Evaluation",
    subtitle = "Bible Society UK 'Quiet Revival' Claim Assessment",
    caption = "Conclusion: Evidence quality is poor (Grade D). Claim not supported by data.",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12),
      plot.caption = element_text(size = 11, hjust = 0.5, face = "bold")
    )
  )
```

## Next Steps

Further analyses will examine:

1.  [**Weekly Attendance Claim**](./weekly-attendance-claim.qmd) - Statistical testing of the 7% → 11% increase
2.  [**Question Order Effects**](./question-order-effects.qmd) - Impact of different question formats
3.  [**Demographic Analysis**](./demographic-analysis.qmd) - How population composition changes affect results