---
title: "Demographic Analysis"
subtitle: "Population composition changes and their impact"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Set working directory to project root using here package
if (!require(here, quietly = TRUE)) {
  install.packages("here")
}
# Find project root (directory containing _quarto.yml)
here::i_am("_quarto.yml")
# Set working directory to project root
setwd(here())

library(tidyverse)
library(scales)
library(ggplot2)
```

## Introduction

Between 2018 and 2024, the UK population composition changed significantly due to immigration, particularly:

- **Ukrainian refugees**: ~217,000-255,000 (post-February 2022)
- **Hong Kong BN(O) visa holders**: >163,000 (January 2021 - March 2025)
- Other net migration changes

These groups may have different religious attendance patterns than the existing UK population, which could explain apparent changes in church attendance.

## Attendance by Ethnicity (2024)

```{r}
#| label: ethnicity-breakdown

attendance_data <- read_csv(here::here("data/bible-society-uk-revival/processed/church-attendance-extracted.csv"))

# Extract 2024 attendance by ethnicity
ethnicity_2024 <- attendance_data %>%
  filter(year == 2024, question_type == "binary", 
         response_category == "Yes - in the past year") %>%
  select(response_category, white, ethnic_minority) %>%
  rename(White = white, `Ethnic Minority` = ethnic_minority)
```

### 2024 Attendance by Ethnicity

- **White**: `r sprintf("%.1f", ethnicity_2024$White)`% attended in past year
- **Ethnic Minority**: `r sprintf("%.1f", ethnicity_2024[['Ethnic Minority']])`% attended in past year

⚠️ **Note**: These figures suggest different attendance patterns between ethnic groups. Population composition changes could significantly affect overall attendance rates.

## Age-Stratified Analysis

```{r}
#| label: age-breakdown

# Weekly attendance for 2018
age_2018 <- attendance_data %>%
  filter(year == 2018, response_category == "At least once a week")

# Weekly attendance for 2024 (sum of categories)
age_2024 <- attendance_data %>%
  filter(
    year == 2024,
    question_type == "frequency",
    response_category %in% c("Daily/almost daily", "A few times a week", "About once a week")
  ) %>%
  group_by(year, question_type) %>%
  summarise(
    across(c(age_18_34, age_35_54, age_55plus), sum, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(response_category = "At least once a week") # Add this for consistency

# Combine and create age_comparison
age_comparison <- bind_rows(age_2018, age_2024) %>%
  select(year, age_18_34, age_35_54, age_55plus) %>%
  rename(
    Year = year,
    `18-34` = age_18_34,
    `35-54` = age_35_54,
    `55+` = age_55plus
  )
```

### 2018

- **18-34**: `r sprintf("%.1f", age_comparison[['18-34']][age_comparison$Year == 2018])`%
- **35-54**: `r sprintf("%.1f", age_comparison[['35-54']][age_comparison$Year == 2018])`%
- **55+**: `r sprintf("%.1f", age_comparison[['55+']][age_comparison$Year == 2018])`%

### 2024

- **18-34**: `r sprintf("%.1f", age_comparison[['18-34']][age_comparison$Year == 2024])`%
- **35-54**: `r sprintf("%.1f", age_comparison[['35-54']][age_comparison$Year == 2024])`%
- **55+**: `r sprintf("%.1f", age_comparison[['55+']][age_comparison$Year == 2024])`%

### Changes (2024 - 2018)

- **18-34**: `r sprintf("%+.1f", age_comparison[['18-34']][age_comparison$Year == 2024] - age_comparison[['18-34']][age_comparison$Year == 2018])` percentage points
- **35-54**: `r sprintf("%+.1f", age_comparison[['35-54']][age_comparison$Year == 2024] - age_comparison[['35-54']][age_comparison$Year == 2018])` percentage points
- **55+**: `r sprintf("%+.1f", age_comparison[['55+']][age_comparison$Year == 2024] - age_comparison[['55+']][age_comparison$Year == 2018])` percentage points

## Visualisation: Age-Stratified Changes

```{r}
#| label: age-visualisation
#| fig-cap: "Change in weekly attendance by age group: 2018 vs 2024"

# Calculate changes in a robust way
data_2018 <- age_comparison %>%
  filter(Year == 2018) %>%
  pivot_longer(cols = -Year, names_to = "Age_Group", values_to = "y2018") %>%
  select(-Year)

data_2024 <- age_comparison %>%
  filter(Year == 2024) %>%
  pivot_longer(cols = -Year, names_to = "Age_Group", values_to = "y2024") %>%
  select(-Year)

age_changes <- full_join(data_2018, data_2024, by = "Age_Group") %>%
  mutate(
    y2018 = ifelse(is.na(y2018), 0, y2018),
    y2024 = ifelse(is.na(y2024), 0, y2024),
    change = y2024 - y2018
  ) %>%
  mutate(
    Age_Group = factor(Age_Group, levels = c("18-34", "35-54", "55+")),
    change_color = ifelse(change > 0, "darkgreen", "darkred")
  )

# Plot the changes
ggplot(age_changes, aes(x = Age_Group, y = change, fill = change_color)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = sprintf("%+.1f", .data$change)), vjust = -0.5) +
  scale_fill_identity() +
  labs(
    x = "Age Group",
    y = "Percentage Point Change",
    title = "Change in Weekly Church Attendance (2018 to 2024)",
    subtitle = "By age group"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## The Demographic Composition Problem

**Critical Issue**: The surveys do not appear to control for demographic composition changes.

### Major population changes 2018-2024

1. **Ukrainian refugees**: 217,000-255,000
   - Higher Orthodox Christian attendance rates
   - Different cultural patterns

2. **Hong Kong BN(O) visa holders**: >163,000
   - Different religious demographics
   - May have higher church attendance rates

3. **Other net migration**: Substantial

### ⚠️ PROBLEM

If the surveys are not weighted or standardised to account for these demographic shifts, any apparent change in attendance could simply reflect:

- More religious immigrants entering the population
- Different attendance patterns of new arrivals
- NOT a 'revival' among existing UK residents

This is a form of 'composition change' vs 'behaviour change' confound that must be addressed to make valid comparisons.

## Implications

Without demographic standardisation, we cannot determine whether:

1. Attendance genuinely increased among existing UK residents (revival)
2. Attendance increased due to immigration (composition effect)
3. A combination of both

**Recommendation**: Any claims about a "revival" should control for demographic composition changes through:
- Demographic weighting adjustments
- Stratified analysis by immigration status
- Age/ethnicity standardisation

Until such analyses are conducted, the demographic composition issue remains a critical red flag.