---
title: "Exploratory Overview"
subtitle: "Data loading, basic summaries, and red flag detection"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Set working directory to project root using here package
if (!require(here, quietly = TRUE)) {
  install.packages("here")
}
# Find project root (directory containing _quarto.yml)
here::i_am("_quarto.yml")
# Set working directory to project root
setwd(here())

# Load required packages
library(tidyverse)
library(scales)
library(ggplot2)
library(knitr)

# Set seed for reproducibility
set.seed(123)
```

## Introduction

This exploratory analysis examines Bible Society UK's claim of a "Quiet Revival" based on YouGov survey data comparing church attendance patterns between 2018 and 2024. We begin by loading the data and conducting initial descriptive analyses and consistency checks.

## Data Loading

```{r}
#| label: load-data

# Load survey metadata
survey_meta <- read_csv(here::here("data/bible-society-uk-revival/processed/survey-metadata.csv"), comment = "#")

# Load attendance data
attendance_data <- read_csv(here::here("data/bible-society-uk-revival/processed/church-attendance-extracted.csv"))

# Display survey metadata
kable(survey_meta, caption = "Survey metadata for 2018 and 2024 surveys")
```

```{r}
#| label: inspect-data
#| include: false

# Quick inspection of data structure
glimpse(attendance_data)
```

## Survey Overview

```{r}
#| label: survey-summary

cat("Survey Comparison:\n")
cat("==================\n\n")
cat(sprintf("2018: n = %s (weighted: %s)\n", 
            format(survey_meta$sample_size_unweighted[1], big.mark = ","),
            format(survey_meta$sample_size_weighted[1], big.mark = ",")))
cat(sprintf("2024: n = %s (weighted: %s)\n", 
            format(survey_meta$sample_size_unweighted[2], big.mark = ","),
            format(survey_meta$sample_size_weighted[2], big.mark = ",")))
cat(sprintf("\nSample size reduction: %.1f%%\n", 
            (1 - survey_meta$sample_size_weighted[2] / survey_meta$sample_size_weighted[1]) * 100))
```

## Red Flag #1: Internal Consistency Check

The 2024 survey asked both a binary question ("Yes - in the past year") and a frequency question. These should yield consistent results, but we need to check.

```{r}
#| label: internal-consistency

# Calculate sum of frequency responses for past year attendance in 2024
freq_2024 <- attendance_data %>%
  filter(year == 2024, question_type == "frequency") %>%
  filter(response_category %in% c(
    "Daily/almost daily", "A few times a week", 
    "About once a week", "About once a fortnight",
    "About once a month", "A few times a year",
    "About once a year"
  ))

freq_sum_2024 <- sum(freq_2024$total_pct, na.rm = TRUE)

# Get binary response
binary_2024 <- attendance_data %>%
  filter(year == 2024, question_type == "binary", 
         response_category == "Yes - in the past year")

binary_pct_2024 <- binary_2024$total_pct

# Display comparison
cat("Internal Consistency Check (2024 Survey):\n")
cat("==========================================\n\n")
cat(sprintf("Binary 'Yes - in the past year': %.1f%%\n", binary_pct_2024))
cat(sprintf("Sum of frequency categories:     %.1f%%\n", freq_sum_2024))
cat(sprintf("Discrepancy:                     %.1f percentage points\n\n", 
            abs(freq_sum_2024 - binary_pct_2024)))

if (abs(freq_sum_2024 - binary_pct_2024) > 2) {
  cat("🚩 RED FLAG: Internal inconsistency detected!\n")
  cat("This suggests measurement error or question order effects.\n")
} else {
  cat("✓ Internal consistency check passed.\n")
}
```

## Red Flag #2: Question Format Differences

The surveys used different question formats, which makes direct comparison problematic:

```{r}
#| label: question-format

cat("Question Format Comparison:\n")
cat("===========================\n\n")
cat("2018: Single frequency question only\n")
cat("     'Apart from weddings, baptisms/christenings, and funerals how often, if at all, did you go to a church service in the last year?'\n\n")
cat("2024: Binary question FIRST, then frequency question\n")
cat("     This order may prime respondents differently.\n\n")
cat("🚩 RED FLAG: Question order effects are a known source of bias.\n")
cat("The binary question may influence responses to the frequency question.\n")
```

## Basic Attendance Statistics

```{r}
#| label: basic-stats

# Calculate key attendance metrics
weekly_2018 <- attendance_data %>%
  filter(year == 2018, response_category == "At least once a week") %>%
  pull(total_pct)

weekly_2024 <- attendance_data %>%
  filter(year == 2024, question_type == "frequency", 
         response_category == "At least once a week") %>%
  pull(total_pct)

never_2018 <- attendance_data %>%
  filter(year == 2018, response_category == "Never") %>%
  pull(total_pct)

never_2024 <- attendance_data %>%
  filter(year == 2024, question_type == "frequency", 
         response_category == "Never") %>%
  pull(total_pct)

# Create summary table
summary_stats <- tibble(
  Metric = c("At least once a week", "Never attended"),
  `2018 (%)` = c(weekly_2018, never_2018),
  `2024 (%)` = c(weekly_2024, never_2024),
  `Change (pp)` = c(weekly_2024 - weekly_2018, never_2024 - never_2018)
)

kable(summary_stats, digits = 1, caption = "Key attendance metrics comparison")
```

## The Contradiction

Notice the contradiction: the weekly attendance measure shows an **increase** (7% → 11%), while the binary "past year" measure shows a **decrease** (\~27% → 24%). These cannot both be true, which suggests measurement error or question order effects.

## Next Steps

Further analyses will examine:

1.  [**Weekly Attendance Claim**](./weekly-attendance-claim.qmd) - Statistical testing of the 7% → 11% increase
2.  [**Question Order Effects**](./question-order-effects.qmd) - Impact of different question formats
3.  [**Demographic Analysis**](./demographic-analysis.qmd) - How population composition changes affect results