---
title: "General Overview"
subtitle: "Comprehensive comparison of 2018 and 2024 survey responses"
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

# Set working directory to project root using here package
if (!require(here, quietly = TRUE)) {
  install.packages("here")
}
# Find project root (directory containing _quarto.yml)
here::i_am("_quarto.yml")
# Set working directory to project root
setwd(here())

# Load required packages
library(tidyverse)
library(scales)
library(ggplot2)
library(knitr)
library(patchwork)

# Set seed for reproducibility
set.seed(123)

# Custom theme
theme_comparison <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(size = 11, color = "grey40"),
      legend.position = "top",
      panel.grid.minor = element_blank()
    )
}
```

## Introduction

This document provides a comprehensive, descriptive comparison of all survey questions common to both the 2018 and 2024 YouGov surveys on church attendance in the UK. The focus here is on presenting the data clearly, showing what was asked and how responses differed between the two time points.

**Important context**: This analysis presents the survey data as reported. However, when compared with actual church attendance records, significant discrepancies emerge. See [Survey Claims vs Church Records](./cofe-comparison.qmd) for a detailed comparison showing that:

- Survey claims 6.66 million weekly attenders
- Actual church records show ~1.46 million across all Christian denominations  
- This 3.7x over-reporting undermines claims based solely on survey data

For immigration context that may explain apparent changes, see [Demographic Analysis](./demographic-analysis.qmd).

## Survey Metadata

```{r}
#| label: load-metadata

# Load survey metadata
survey_meta <- read_csv(here::here("data/bible-society-uk-revival/processed/survey-metadata.csv"), 
                        comment = "#")
```

::: {style="overflow-x: auto; overflow-y: auto; max-height: 400px; border: 1px solid #ddd; padding: 10px; margin-bottom: 20px;"}

```{r}
#| echo: false

kable(survey_meta, caption = "Survey metadata for 2018 and 2024")
```

:::

### Key Survey Characteristics

```{r}
#| label: survey-characteristics-table

characteristics <- tibble(
  Characteristic = c(
    "Survey Organisation",
    "Survey Type",
    "Fieldwork Period",
    "Sample Size (Unweighted)",
    "Sample Size (Weighted)",
    "Primary Question Topic"
  ),
  `2018` = c(
    "YouGov",
    "Online panel",
    "11 Oct - 13 Nov 2018",
    format(survey_meta$sample_size_unweighted[1], big.mark = ","),
    format(survey_meta$sample_size_weighted[1], big.mark = ","),
    "Church attendance frequency"
  ),
  `2024` = c(
    "YouGov",
    "Online panel",
    "4 Nov - 2 Dec 2024",
    format(survey_meta$sample_size_unweighted[2], big.mark = ","),
    format(survey_meta$sample_size_weighted[2], big.mark = ","),
    "Church attendance (binary + frequency)"
  )
)

kable(characteristics, caption = "Survey characteristics comparison")
```

**Note**: The 2024 survey has `r sprintf("%.1f", (1 - survey_meta$sample_size_weighted[2] / survey_meta$sample_size_weighted[1]) * 100)`% fewer respondents than 2018.

### Sample Size and Statistical Precision

```{r}
#| label: sample-size-precision
#| echo: false

# Calculate margins of error for both surveys
n_2018 <- survey_meta$sample_size_weighted[1]
n_2024 <- survey_meta$sample_size_weighted[2]

# Calculate margin of error for 50% proportion (worst case scenario) at 95% confidence
moe_2018 <- 1.96 * sqrt(0.5 * 0.5 / n_2018) * 100
moe_2024 <- 1.96 * sqrt(0.5 * 0.5 / n_2024) * 100

# Calculate effective sample sizes assuming design effect of 1.5 (typical for online panels)
deff <- 1.5
n_eff_2018 <- n_2018 / deff
n_eff_2024 <- n_2024 / deff
moe_eff_2018 <- 1.96 * sqrt(0.5 * 0.5 / n_eff_2018) * 100
moe_eff_2024 <- 1.96 * sqrt(0.5 * 0.5 / n_eff_2024) * 100
```

**Central Limit Theorem Implications:**

According to the Central Limit Theorem, with sufficiently large sample sizes, the sampling distribution of proportions approaches a normal distribution, allowing us to make reliable statistical inferences. Both surveys have sample sizes well above the typical threshold (n ≥ 30) required for the CLT to apply.

**Statistical Precision Comparison:**

| Survey | Sample Size | Margin of Error (95% CI) | With Design Effect (1.5) |
|--------|-------------|--------------------------|--------------------------|
| 2018 | `r format(n_2018, big.mark = ",")` | ±`r sprintf("%.2f", moe_2018)`% | ±`r sprintf("%.2f", moe_eff_2018)`% |
| 2024 | `r format(n_2024, big.mark = ",")` | ±`r sprintf("%.2f", moe_2024)`% | ±`r sprintf("%.2f", moe_eff_2024)`% |

**Key Points:**

1. **Both samples are statistically adequate**: With n > 12,000 in both surveys, the Central Limit Theorem ensures the sampling distributions are approximately normal, providing robust statistical properties.

2. **Minimal precision loss**: The 2024 survey has a margin of error only ~`r sprintf("%.2f", moe_2024 - moe_2018)` percentage points wider than 2018. This difference is negligible for practical purposes.

3. **Design effect adjustment**: Assuming a design effect of 1.5 (typical for online panel surveys), effective sample sizes are still large (`r format(round(n_eff_2018), big.mark = ",")` and `r format(round(n_eff_2024), big.mark = ",")` respectively), with margins of error remaining under 1%.

4. **Representativeness vs. Precision**: The more important concern is not statistical precision (both surveys are precise) but **representativeness**:
   - **Why did the sample size drop 31%?** Potential reasons include:
     - Changes in response rates
     - Different sampling frame or recruitment methods
     - Changes in panel composition
     - Different fielding conditions
   
5. **Selection bias risk**: The substantial reduction in sample size may indicate differential non-response or systematic changes in who participates in the survey. This could affect representativeness even though statistical precision remains high.

**Conclusion**: From a pure Central Limit Theorem perspective, both sample sizes are excellent and provide similar statistical precision. However, the unexplained 31% reduction raises questions about potential changes in survey methodology or response patterns that could affect comparability between the two time points.

## Data Loading

```{r}
#| label: load-data

# Load attendance data
attendance_data <- read_csv(here::here("data/bible-society-uk-revival/processed/church-attendance-extracted.csv"))
```

## Survey Question Overview

This section examines how many questions about church attendance were asked in each survey and which questions are comparable.

```{r}
#| label: question-overview
#| echo: false

# Count unique question types per survey
questions_2018 <- attendance_data %>%
  filter(year == 2018) %>%
  distinct(question_type) %>%
  nrow()

questions_2024 <- attendance_data %>%
  filter(year == 2024) %>%
  distinct(question_type) %>%
  nrow()

# Identify common question types
common_questions <- attendance_data %>%
  group_by(question_type) %>%
  summarise(years = n_distinct(year)) %>%
  filter(years == 2) %>%
  nrow()
```

### Question Count Summary

| Survey Year | Number of Church Attendance Questions | Question Types |
|-------------|--------------------------------------|----------------|
| **2018** | `r questions_2018` | Frequency only |
| **2024** | `r questions_2024` | Binary + Frequency |
| **Common Questions** | `r common_questions` | Frequency question |

### Detailed Breakdown

**2018 Survey (1 question):**

1. **Frequency Question**: "Apart from weddings, baptisms/christenings, and funerals how often, if at all, did you go to a church service in the last year?"
   - Response scale: 9 categories from "Daily/almost daily" to "Never"

**2024 Survey:**

1. **Frequency Question** (SAME as 2018): "Apart from weddings, baptisms/christenings, and funerals how often, if at all, did you go to a church service in the last year?"
   - Response scale: Same 9 categories as 2018
   - **Note**: There is also a separate question about Bible engagement at church services (part of "Have you read, listened to, or engaged with the Bible in any of the following locations/situations?"), but this measures a different construct than attendance.

### Comparability Assessment

```{r}
#| label: comparability-visual
#| echo: false
#| fig-cap: "Survey question structure comparison"
#| fig-width: 10
#| fig-height: 4

# Create visualization of question structure
question_structure <- tibble(
  Survey = c("2018", "2024", "2024"),
  Question = c("Frequency Question", "Binary Question (NEW)", "Frequency Question"),
  Order = c(1, 1, 2),
  Comparable = c("Yes", "No", "Yes"),
  y_pos = c(1.5, 2, 1),
  x_pos = c(1, 1, 1)
)

ggplot(question_structure, aes(x = x_pos, y = y_pos)) +
  geom_tile(aes(fill = Comparable), color = "white", linewidth = 2, 
            width = 0.8, height = 0.4, alpha = 0.8) +
  geom_text(aes(label = Question), fontface = "bold", size = 3.5, 
            lineheight = 0.9) +
  geom_text(aes(label = paste0("Order: ", Order)), nudge_y = -0.15, 
            size = 3, color = "grey30") +
  facet_wrap(~Survey, ncol = 2) +
  scale_fill_manual(values = c("Yes" = "#41ab5d", "No" = "#d73027")) +
  labs(
    title = "Survey Question Structure: 2018 vs 2024",
    subtitle = "Green = Directly comparable | Red = Not comparable (unique to 2024)",
    fill = "Comparable?"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5, color = "grey40"),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "bottom",
    plot.margin = margin(10, 10, 10, 10)
  )
```

### Key Observations

1. **Direct Comparability**: Both surveys use the same frequency question about church attendance, making them directly comparable.

2. **Additional Question in 2024**: The 2024 survey includes a separate question about Bible engagement at church services (part of a question about Bible engagement in different locations). This measures a different construct (Bible engagement) than attendance, so it should not be compared directly to attendance figures.

3. **No Question Order Effect**: Since there is no binary attendance question asked before the frequency question in 2024, there is no question order effect affecting the comparability of the attendance frequency question.

**Conclusion**: Both surveys use the same frequency question about church attendance with identical wording and response categories. The questions are directly comparable. The 2024 survey's additional Bible engagement question measures a different construct and should not be confused with attendance.

## Complete Survey Question Inventory

This section provides a comprehensive inventory of ALL questions asked in both the 2018 and 2024 surveys, showing which questions are common and which are unique to each survey.

### Question Categories

Based on the survey documentation, both surveys covered multiple topics related to religion and church attendance. Below is a complete comparison:

```{r}
#| label: complete-question-inventory
#| echo: false

# Create comprehensive question inventory
question_inventory <- tibble(
  Category = c(
    # Church Attendance
    "Church Attendance", "Church Attendance", "Church Attendance",
    # Religious Identification
    "Religious Identification", "Religious Identification",
    # Bible and Religious Beliefs
    "Bible/Religious Beliefs", "Bible/Religious Beliefs", "Bible/Religious Beliefs",
    # Demographics
    "Demographics", "Demographics", "Demographics"
  ),
  Question_Topic = c(
    # Church Attendance
    "Frequency of church service attendance (past year)",
    "Bible engagement at church services (2024 only, different construct)",
    "When last attended (if not in past year)",
    # Religious Identification
    "Religious affiliation/denomination",
    "Christian denomination (if Christian)",
    # Bible and Religious Beliefs
    "Bible reading frequency",
    "Bible can change society for the better",
    "Other religious belief questions",
    # Demographics
    "Age group",
    "Ethnicity",
    "Gender"
  ),
  Asked_2018 = c(
    "Yes", "No", "No",
    "Yes", "Yes",
    "Yes", "Yes", "Yes",
    "Yes", "Limited", "Limited"
  ),
  Asked_2024 = c(
    "Yes", "Yes", "Yes",
    "Yes", "Yes",
    "Yes", "Yes", "Yes",
    "Yes", "Yes", "Yes"
  ),
  Directly_Comparable = c(
    "Yes", "No (different construct)", "No",
    "Yes", "Yes",
    "Unknown**", "Unknown**", "Unknown**",
    "Yes", "Partial***", "Partial***"
  ),
  Data_Available = c(
    "Yes", "Yes (2024 only)", "Yes (2024 only)",
    "No", "No",
    "No", "No", "No",
    "Yes", "Yes (2024 only)", "Yes (2024 only)"
  )
)

kable(question_inventory, 
      col.names = c("Category", "Question Topic", "2018", "2024", "Comparable?", "Data Available"),
      caption = "Complete inventory of survey questions across both years")
```

**Notes:**

- \* **Full comparability**: Question wording identical, same context
- \*\* **Unknown comparability**: Questions asked in both surveys, but exact wording not documented in our data extracts
- \*\*\* **Partial demographic data**: 2018 had limited ethnicity/gender breakdowns compared to 2024

### Summary Statistics

```{r}
#| label: question-summary-stats
#| echo: false

# Calculate summary statistics
total_topics_2018 <- question_inventory %>% filter(Asked_2018 == "Yes") %>% nrow()
total_topics_2024 <- question_inventory %>% filter(Asked_2024 == "Yes") %>% nrow()
common_topics <- question_inventory %>% 
  filter(Asked_2018 == "Yes" & Asked_2024 == "Yes") %>% 
  nrow()
unique_2024 <- question_inventory %>% 
  filter(Asked_2018 == "No" & Asked_2024 == "Yes") %>% 
  nrow()
comparable_topics <- question_inventory %>% 
  filter(Directly_Comparable == "Yes") %>% 
  nrow()
```

| Metric | Count |
|--------|-------|
| **Topics covered in 2018** | `r total_topics_2018` |
| **Topics covered in 2024** | `r total_topics_2024` |
| **Common topics (asked in both)** | `r common_topics` |
| **Unique to 2024** | `r unique_2024` |
| **Directly comparable** | `r comparable_topics` |
| **Data extracted and available for analysis** | 4 topics |

### Visual Summary: Question Coverage

```{r}
#| label: question-coverage-visual
#| fig-cap: "Venn diagram style comparison of survey question coverage"
#| fig-width: 10
#| fig-height: 6
#| echo: false

# Create a visual representation of question overlap
coverage_data <- tibble(
  Survey = c(rep("2018", total_topics_2018), rep("2024", total_topics_2024)),
  Topic_Number = c(1:total_topics_2018, 1:total_topics_2024),
  Category = c(
    question_inventory %>% filter(Asked_2018 == "Yes") %>% pull(Category),
    question_inventory %>% filter(Asked_2024 == "Yes") %>% pull(Category)
  ),
  Status = c(
    question_inventory %>% filter(Asked_2018 == "Yes") %>% 
      mutate(Status = ifelse(Asked_2024 == "Yes", "Common", "2018 Only")) %>% 
      pull(Status),
    question_inventory %>% filter(Asked_2024 == "Yes") %>% 
      mutate(Status = ifelse(Asked_2018 == "Yes", "Common", "2024 Only")) %>% 
      pull(Status)
  )
)

ggplot(coverage_data, aes(x = Survey, fill = Status)) +
  geom_bar(position = "stack", alpha = 0.85, width = 0.6) +
  geom_text(stat = "count", aes(label = after_stat(count)), 
            position = position_stack(vjust = 0.5), 
            fontface = "bold", size = 6, color = "white") +
  scale_fill_manual(values = c("Common" = "#41ab5d", 
                                "2018 Only" = "#4575b4",
                                "2024 Only" = "#d73027")) +
  labs(
    title = "Survey Question Coverage: 2018 vs 2024",
    subtitle = "Number of question topics by survey year",
    x = "Survey Year",
    y = "Number of Question Topics",
    fill = "Status"
  ) +
  theme_comparison() +
  theme(legend.position = "right")
```

### Key Findings from Complete Inventory

1. **Both surveys covered similar topic areas**: Church attendance, religious identification, beliefs about the Bible, and demographics

2. **2024 includes additional question**: Added question about Bible engagement at church services (part of a question about Bible engagement in different locations), which measures a different construct than attendance

3. **Demographic data improved in 2024**: More comprehensive ethnicity and gender breakdowns

4. **Limited data extraction**: Our analysis focuses on church attendance patterns because:
   - These are the questions highlighted in Bible Society UK's "Quiet Revival" claim
   - Full cross-tabulations for other questions were not included in the published reports
   - Demographic data was only partially available for 2018

5. **Comparability challenges**: Even for topics covered in both years:
   - Question wording may have changed (not fully documented)
   - Response scales may differ
   - Question context/ordering changed
   - Demographic weighting methodology not disclosed

### Data Availability for Analysis

Of all the questions asked across both surveys, we have extracted and analyzable data for:

- ✅ **Church attendance frequency** (both years)
- ✅ **Binary church attendance** (2024 only)
- ✅ **Age breakdowns** (both years)
- ✅ **Ethnicity breakdowns** (2024 only, limited in 2018)
- ✅ **Gender breakdowns** (2024 only)

For other questions (religious identification, Bible reading, beliefs), the published reports provided:
- Summary statistics only (no cross-tabulations)
- Selected highlights (not comprehensive data)
- Insufficient detail for rigorous comparative analysis

This limitation means our analysis necessarily focuses on what data is available: church attendance patterns and basic demographic breakdowns.

## Question Format Comparison

### 2018 Survey Question

**Question wording**: "Apart from weddings, baptisms/christenings, and funerals how often, if at all, did you go to a church service in the last year?"

**Response options**:
- Daily/almost daily
- A few times a week
- About once a week
- About once a fortnight
- About once a month
- A few times a year
- About once a year
- Hardly ever
- Never

### 2024 Survey Questions

**Two separate questions were asked:**

**Question 1 (Binary)**: "Have you attended a church service?" [exact wording not fully documented]

**Response options**:
- Yes - in the past year
- Yes - more than a year ago
- Never

**Question 2 (Frequency)**: Same frequency scale as 2018

⚠️ **Important methodological note**: Both surveys use the same frequency question. The 2024 survey also includes a separate question about Bible engagement at church services, but this measures a different construct than attendance.

## Complete Attendance Distribution Comparison

### Frequency Question Responses

This is the most directly comparable question between the two surveys.

```{r}
#| label: frequency-distribution-table

# Extract frequency responses for both years
freq_2018 <- attendance_data %>%
  filter(year == 2018, 
         !response_category %in% c("At least once a week", "At least once a month")) %>%
  select(response_category, total_pct) %>%
  rename(`2018 (%)` = total_pct)

freq_2024 <- attendance_data %>%
  filter(year == 2024, question_type == "frequency",
         !response_category %in% c("At least once a week", "At least once a month")) %>%
  select(response_category, total_pct) %>%
  rename(`2024 (%)` = total_pct)

# Merge and calculate changes
freq_comparison <- full_join(freq_2018, freq_2024, by = "response_category") %>%
  mutate(
    `Change (pp)` = `2024 (%)` - `2018 (%)`,
    `Change (%)` = round((`2024 (%)` / `2018 (%)` - 1) * 100, 1)
  ) %>%
  arrange(desc(`2018 (%)`))

kable(freq_comparison, digits = 1, 
      caption = "Church attendance frequency: 2018 vs 2024")
```

### Visual Comparison: Complete Distribution

```{r}
#| label: full-distribution-visual
#| fig-cap: "Complete church attendance distribution comparison"
#| fig-width: 12
#| fig-height: 7

# Prepare data for plotting
plot_data <- attendance_data %>%
  filter(
    (year == 2018) | (year == 2024 & question_type == "frequency")
  ) %>%
  filter(!response_category %in% c("At least once a week", "At least once a month")) %>%
  mutate(
    Year = factor(year),
    response_category = factor(response_category, levels = c(
      "Daily/almost daily", "A few times a week", "About once a week",
      "About once a fortnight", "About once a month", "A few times a year",
      "About once a year", "Hardly ever", "Never"
    ))
  )

ggplot(plot_data, aes(x = response_category, y = total_pct, fill = Year)) +
  geom_col(position = "dodge", alpha = 0.8, width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", total_pct)), 
            position = position_dodge(width = 0.7), 
            vjust = -0.3, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("2018" = "#4575b4", "2024" = "#d73027"),
                    labels = c("2018" = "2018 (n=19,875)", "2024" = "2024 (n=12,455)")) +
  labs(
    title = "Church Attendance Frequency Distribution: 2018 vs 2024",
    subtitle = "Direct comparison of identical question format",
    x = "Attendance Frequency",
    y = "Percentage of Respondents (%)",
    fill = "Survey Year"
  ) +
  theme_comparison() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  ) +
  scale_y_continuous(limits = c(0, 70), breaks = seq(0, 70, 10))
```

### Change Magnitude Visualisation

```{r}
#| label: change-magnitude-visual
#| fig-cap: "Magnitude and direction of changes in attendance categories"
#| fig-width: 10
#| fig-height: 6

# Prepare data for diverging bar chart
change_plot_data <- freq_comparison %>%
  mutate(
    response_category = factor(response_category, 
                               levels = rev(response_category)),
    Direction = case_when(
      `Change (pp)` > 0 ~ "Increase",
      `Change (pp)` < 0 ~ "Decrease",
      TRUE ~ "No Change"
    )
  )

ggplot(change_plot_data, aes(x = response_category, y = `Change (pp)`, fill = Direction)) +
  geom_col(alpha = 0.85, width = 0.7) +
  geom_text(aes(label = sprintf("%+.1f", `Change (pp)`)), 
            hjust = ifelse(change_plot_data$`Change (pp)` > 0, -0.2, 1.2),
            size = 4, fontface = "bold") +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.8) +
  scale_fill_manual(values = c("Increase" = "#2166ac", "Decrease" = "#d73027")) +
  coord_flip() +
  labs(
    title = "Change in Attendance by Category: 2018 → 2024",
    subtitle = "Percentage point change for each response category",
    x = NULL,
    y = "Change (percentage points)",
    fill = "Direction"
  ) +
  theme_comparison() +
  theme(
    panel.grid.major.y = element_blank()
  )
```

## Summary Statistics: Grouped Categories

### Weekly and Monthly Attendance

```{r}
#| label: grouped-categories

# Calculate grouped statistics
grouped_stats <- tibble(
  Category = c("At least once a week", "At least once a month", "Never"),
  `2018 (%)` = c(
    attendance_data %>% filter(year == 2018, response_category == "At least once a week") %>% pull(total_pct),
    attendance_data %>% filter(year == 2018, response_category == "At least once a month") %>% pull(total_pct),
    attendance_data %>% filter(year == 2018, response_category == "Never") %>% pull(total_pct)
  ),
  `2024 (%)` = c(
    # Weekly: sum of daily + few times a week + once a week
    attendance_data %>% 
      filter(year == 2024, question_type == "frequency",
             response_category %in% c("Daily/almost daily", "A few times a week", "About once a week")) %>%
      pull(total_pct) %>% sum(),
    # Monthly: weekly + fortnight + month
    attendance_data %>% 
      filter(year == 2024, question_type == "frequency",
             response_category %in% c("Daily/almost daily", "A few times a week", "About once a week",
                                     "About once a fortnight", "About once a month")) %>%
      pull(total_pct) %>% sum(),
    # Never
    attendance_data %>% filter(year == 2024, question_type == "frequency", response_category == "Never") %>% pull(total_pct)
  )
) %>%
  mutate(
    `Change (pp)` = `2024 (%)` - `2018 (%)`,
    `Relative Change (%)` = round((`2024 (%)` / `2018 (%)` - 1) * 100, 1)
  )

kable(grouped_stats, digits = 1, 
      caption = "Grouped attendance categories comparison")
```

```{r}
#| label: grouped-visual
#| fig-cap: "Comparison of grouped attendance categories"
#| fig-width: 10
#| fig-height: 5

# Reshape for plotting
grouped_plot_data <- grouped_stats %>%
  select(Category, `2018 (%)`, `2024 (%)`) %>%
  pivot_longer(cols = c(`2018 (%)`, `2024 (%)`), 
               names_to = "Year", values_to = "Percentage") %>%
  mutate(
    Year = str_extract(Year, "\\d{4}"),
    Category = factor(Category, levels = c("At least once a week", "At least once a month", "Never"))
  )

ggplot(grouped_plot_data, aes(x = Category, y = Percentage, fill = Year)) +
  geom_col(position = "dodge", alpha = 0.85, width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", Percentage)), 
            position = position_dodge(width = 0.7), 
            vjust = -0.3, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("2018" = "#4575b4", "2024" = "#d73027")) +
  labs(
    title = "Grouped Attendance Categories: 2018 vs 2024",
    subtitle = "Weekly and monthly attendance show increases; 'Never' shows decrease",
    x = NULL,
    y = "Percentage (%)",
    fill = "Survey Year"
  ) +
  theme_comparison() +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1)
  ) +
  scale_y_continuous(limits = c(0, 70), breaks = seq(0, 70, 10))
```

## Demographic Breakdowns

### By Age Group

```{r}
#| label: age-breakdown-table

# Get weekly attendance by age for 2018
age_2018 <- attendance_data %>%
  filter(year == 2018, response_category == "At least once a week") %>%
  select(age_18_34, age_35_54, age_55plus)

# Get weekly attendance by age for 2024
age_2024 <- attendance_data %>%
  filter(
    year == 2024,
    question_type == "frequency",
    response_category %in% c("Daily/almost daily", "A few times a week", "About once a week")
  ) %>%
  summarise(
    age_18_34 = sum(age_18_34, na.rm = TRUE),
    age_35_54 = sum(age_35_54, na.rm = TRUE),
    age_55plus = sum(age_55plus, na.rm = TRUE)
  )

# Create comparison table
age_comparison <- tibble(
  `Age Group` = c("18-34", "35-54", "55+"),
  `2018 (%)` = c(age_2018$age_18_34, age_2018$age_35_54, age_2018$age_55plus),
  `2024 (%)` = c(age_2024$age_18_34, age_2024$age_35_54, age_2024$age_55plus)
) %>%
  mutate(
    `Change (pp)` = `2024 (%)` - `2018 (%)`,
    `Relative Change (%)` = round((`2024 (%)` / `2018 (%)` - 1) * 100, 1)
  )

kable(age_comparison, digits = 1, 
      caption = "Weekly+ church attendance by age group")
```

```{r}
#| label: age-visual
#| fig-cap: "Weekly+ attendance by age group: 2018 vs 2024"
#| fig-width: 10
#| fig-height: 6

# Reshape for plotting
age_plot_data <- age_comparison %>%
  select(`Age Group`, `2018 (%)`, `2024 (%)`) %>%
  pivot_longer(cols = c(`2018 (%)`, `2024 (%)`), 
               names_to = "Year", values_to = "Percentage") %>%
  mutate(Year = str_extract(Year, "\\d{4}"))

ggplot(age_plot_data, aes(x = `Age Group`, y = Percentage, fill = Year)) +
  geom_col(position = "dodge", alpha = 0.85, width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", Percentage)), 
            position = position_dodge(width = 0.6), 
            vjust = -0.3, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("2018" = "#4575b4", "2024" = "#d73027")) +
  labs(
    title = "Weekly+ Church Attendance by Age Group",
    subtitle = "Comparison between 2018 and 2024",
    x = "Age Group",
    y = "Percentage Attending Weekly or More (%)",
    fill = "Survey Year"
  ) +
  theme_comparison() +
  scale_y_continuous(limits = c(0, 20), breaks = seq(0, 20, 5))
```

### Change by Age Group

```{r}
#| label: age-change-visual
#| fig-cap: "Absolute change in weekly+ attendance by age group"
#| fig-width: 10
#| fig-height: 5

ggplot(age_comparison, aes(x = `Age Group`, y = `Change (pp)`)) +
  geom_col(fill = "#2166ac", alpha = 0.85, width = 0.6) +
  geom_text(aes(label = sprintf("%+.1f pp", `Change (pp)`)), 
            vjust = -0.5, size = 5, fontface = "bold") +
  geom_hline(yintercept = 0, linewidth = 0.8) +
  labs(
    title = "Change in Weekly+ Attendance by Age Group (2018 → 2024)",
    subtitle = "All age groups show increases, with 18-34 showing the largest change",
    x = "Age Group",
    y = "Change (percentage points)"
  ) +
  theme_comparison() +
  scale_y_continuous(limits = c(0, 9), breaks = seq(0, 9, 1), 
                     expand = expansion(mult = c(0, 0.05)))
```

### By Ethnicity (2024 Only)

The 2024 survey includes ethnicity breakdowns that were not available in 2018.

```{r}
#| label: ethnicity-table

# Get 2024 binary attendance by ethnicity
ethnicity_data <- attendance_data %>%
  filter(year == 2024, question_type == "binary", 
         response_category == "Yes - in the past year") %>%
  select(white, ethnic_minority)

ethnicity_table <- tibble(
  `Ethnic Group` = c("White", "Ethnic Minority"),
  `Attended in Past Year (%)` = c(ethnicity_data$white, ethnicity_data$ethnic_minority),
  `Difference from White (pp)` = c(0, ethnicity_data$ethnic_minority - ethnicity_data$white)
)

kable(ethnicity_table, digits = 1, 
      caption = "Church attendance by ethnicity (2024 survey only)")
```

```{r}
#| label: ethnicity-visual
#| fig-cap: "Church attendance by ethnicity (2024)"
#| fig-width: 8
#| fig-height: 5

ggplot(ethnicity_table, aes(x = `Ethnic Group`, y = `Attended in Past Year (%)`)) +
  geom_col(fill = c("#4575b4", "#d73027"), alpha = 0.85, width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", `Attended in Past Year (%)`)), 
            vjust = -0.3, size = 5, fontface = "bold") +
  labs(
    title = "Church Attendance by Ethnicity (2024)",
    subtitle = "Percentage attending in the past year",
    x = "Ethnic Group",
    y = "Attended in Past Year (%)"
  ) +
  theme_comparison() +
  scale_y_continuous(limits = c(0, 30), breaks = seq(0, 30, 5))
```

### By Gender (2024 Only)

```{r}
#| label: gender-table

# Get 2024 binary attendance by gender
gender_data <- attendance_data %>%
  filter(year == 2024, question_type == "binary", 
         response_category == "Yes - in the past year") %>%
  select(male, female)

gender_table <- tibble(
  Gender = c("Male", "Female"),
  `Attended in Past Year (%)` = c(gender_data$male, gender_data$female),
  `Difference from Male (pp)` = c(0, gender_data$female - gender_data$male)
)

kable(gender_table, digits = 1, 
      caption = "Church attendance by gender (2024 survey only)")
```

```{r}
#| label: gender-visual
#| fig-cap: "Church attendance by gender (2024)"
#| fig-width: 8
#| fig-height: 5

ggplot(gender_table, aes(x = Gender, y = `Attended in Past Year (%)`)) +
  geom_col(fill = c("#4575b4", "#d73027"), alpha = 0.85, width = 0.5) +
  geom_text(aes(label = sprintf("%.1f%%", `Attended in Past Year (%)`)), 
            vjust = -0.3, size = 5, fontface = "bold") +
  labs(
    title = "Church Attendance by Gender (2024)",
    subtitle = "Percentage attending in the past year",
    x = "Gender",
    y = "Attended in Past Year (%)"
  ) +
  theme_comparison() +
  scale_y_continuous(limits = c(0, 30), breaks = seq(0, 30, 5))
```

## 2024 Bible Engagement Question Results

The 2024 survey included a question about Bible engagement at church services (part of a question about Bible engagement in different locations). This measures a different construct than attendance.

```{r}
#| label: binary-question-table

binary_data <- attendance_data %>%
  filter(year == 2024, question_type == "binary") %>%
  select(response_category, total_pct) %>%
  rename(`Percentage (%)` = total_pct)

kable(binary_data, digits = 1, 
      caption = "2024 binary question responses")
```

```{r}
#| label: binary-visual
#| fig-cap: "2024 binary question responses"
#| fig-width: 10
#| fig-height: 6

ggplot(binary_data, aes(x = reorder(response_category, -`Percentage (%)`), 
                        y = `Percentage (%)`)) +
  geom_col(fill = c("#d73027", "#fdae61", "#4575b4", "#91bfdb"), 
           alpha = 0.85, width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", `Percentage (%)`)), 
            vjust = -0.3, size = 5, fontface = "bold") +
  labs(
    title = "Binary Question Responses (2024)",
    subtitle = "Have you attended a church service?",
    x = "Response",
    y = "Percentage (%)"
  ) +
  theme_comparison() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)) +
  scale_y_continuous(limits = c(0, 50), breaks = seq(0, 50, 10))
```

## Internal Consistency Check (2024)

We can check whether the binary and frequency questions in 2024 give consistent results.

```{r}
#| label: consistency-check

# Calculate sum of frequency responses for past year attendance in 2024
freq_sum_2024 <- attendance_data %>%
  filter(year == 2024, question_type == "frequency") %>%
  filter(response_category %in% c(
    "Daily/almost daily", "A few times a week", 
    "About once a week", "About once a fortnight",
    "About once a month", "A few times a year",
    "About once a year"
  )) %>%
  pull(total_pct) %>%
  sum()

# Get binary response
binary_2024 <- attendance_data %>%
  filter(year == 2024, question_type == "binary", 
         response_category == "Yes - in the past year") %>%
  pull(total_pct)

consistency <- tibble(
  Measure = c("Binary: 'Yes - in the past year'", 
              "Frequency: Sum of all 'past year' categories",
              "Discrepancy"),
  Value = c(binary_2024, freq_sum_2024, abs(freq_sum_2024 - binary_2024))
)

kable(consistency, digits = 1, col.names = c("Measure", "Percentage (%)"),
      caption = "Internal consistency check: 2024 survey")
```

```{r}
#| label: consistency-visual
#| fig-cap: "Internal consistency comparison"
#| fig-width: 8
#| fig-height: 5

consistency_plot <- tibble(
  Measure = c("Binary Question\n('Yes - in past year')", 
              "Frequency Question\n(Sum of past year categories)"),
  Value = c(binary_2024, freq_sum_2024)
)

ggplot(consistency_plot, aes(x = Measure, y = Value)) +
  geom_col(fill = c("#d73027", "#4575b4"), alpha = 0.85, width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", Value)), 
            vjust = -0.3, size = 6, fontface = "bold") +
  geom_segment(aes(x = 1, xend = 2, y = binary_2024, yend = freq_sum_2024),
               linetype = "dashed", linewidth = 1, color = "grey40") +
  annotate("text", x = 1.5, y = (binary_2024 + freq_sum_2024)/2, 
           label = sprintf("Difference:\n%.1f pp", abs(freq_sum_2024 - binary_2024)),
           size = 4, fontface = "italic") +
  labs(
    title = "Internal Consistency Check: 2024 Survey",
    subtitle = "Do the binary and frequency questions give the same answer?",
    x = NULL,
    y = "Percentage (%)"
  ) +
  theme_comparison() +
  scale_y_continuous(limits = c(0, 30), breaks = seq(0, 30, 5))
```

**Observation**: The two measures differ by `r sprintf("%.1f", abs(freq_sum_2024 - binary_2024))` percentage points. These should theoretically be identical if measuring the same construct.

## Summary Dashboard

```{r}
#| label: summary-dashboard
#| fig-cap: "Summary dashboard of key findings"
#| fig-width: 12
#| fig-height: 10
#| echo: false

# Create summary metrics
p1 <- ggplot(grouped_stats %>% filter(Category == "At least once a week"), 
             aes(x = 1)) +
  geom_segment(aes(x = 1, xend = 1, y = `2018 (%)`, yend = `2024 (%)`),
               arrow = arrow(length = unit(0.3, "cm")), 
               linewidth = 3, color = "#2166ac") +
  geom_point(aes(y = `2018 (%)`), size = 12, color = "#4575b4") +
  geom_point(aes(y = `2024 (%)`), size = 12, color = "#d73027") +
  geom_text(aes(y = `2018 (%)`, label = sprintf("%.0f%%", `2018 (%)`)), 
            color = "white", fontface = "bold", size = 3) +
  geom_text(aes(y = `2024 (%)`, label = sprintf("%.0f%%", `2024 (%)`)), 
            color = "white", fontface = "bold", size = 3) +
  annotate("text", x = 1, y = 14, label = sprintf("+%.1fpp", grouped_stats$`Change (pp)`[1]), 
           fontface = "bold", size = 6, color = "#2166ac") +
  labs(title = "Weekly+ Attendance", subtitle = "2018 → 2024") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13),
        plot.subtitle = element_text(hjust = 0.5, size = 10)) +
  ylim(0, 15)

p2 <- ggplot(grouped_stats %>% filter(Category == "Never"), 
             aes(x = 1)) +
  geom_segment(aes(x = 1, xend = 1, y = `2018 (%)`, yend = `2024 (%)`),
               arrow = arrow(length = unit(0.3, "cm")), 
               linewidth = 3, color = "#a50f15") +
  geom_point(aes(y = `2018 (%)`), size = 12, color = "#4575b4") +
  geom_point(aes(y = `2024 (%)`), size = 12, color = "#d73027") +
  geom_text(aes(y = `2018 (%)`, label = sprintf("%.0f%%", `2018 (%)`)), 
            color = "white", fontface = "bold", size = 3) +
  geom_text(aes(y = `2024 (%)`, label = sprintf("%.0f%%", `2024 (%)`)), 
            color = "white", fontface = "bold", size = 3) +
  annotate("text", x = 1, y = 55, label = sprintf("%.1fpp", grouped_stats$`Change (pp)`[3]), 
           fontface = "bold", size = 6, color = "#a50f15") +
  labs(title = "Never Attended", subtitle = "2018 → 2024") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 13),
        plot.subtitle = element_text(hjust = 0.5, size = 10)) +
  ylim(50, 70)

p3 <- ggplot(age_comparison, aes(x = `Age Group`, y = `Change (pp)`)) +
  geom_col(fill = "#2166ac", alpha = 0.85) +
  geom_text(aes(label = sprintf("%+.1f", `Change (pp)`)), 
            vjust = -0.3, size = 4, fontface = "bold") +
  labs(title = "Change by Age Group", 
       subtitle = "Weekly+ attendance change (pp)",
       x = NULL, y = "Change (pp)") +
  theme_comparison() +
  theme(plot.title = element_text(size = 13))

p4 <- ggplot(consistency_plot, aes(x = Measure, y = Value)) +
  geom_col(fill = c("#d73027", "#4575b4"), alpha = 0.85) +
  geom_text(aes(label = sprintf("%.1f%%", Value)), 
            vjust = -0.3, size = 4, fontface = "bold") +
  labs(title = "2024 Consistency Check",
       subtitle = "Binary vs Frequency",
       x = NULL, y = "%") +
  theme_comparison() +
  theme(
    plot.title = element_text(size = 13),
    axis.text.x = element_text(size = 8)
  )

# Combine plots
(p1 | p2) / (p3 | p4) +
  plot_annotation(
    title = "Key Findings Summary: 2018 vs 2024",
    subtitle = "Comprehensive overview of main attendance pattern changes",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12)
    )
  )
```

## Key Observations

### Main Findings

1. **Weekly+ attendance increased** from 7% (2018) to 11% (2024) - a change of **+4 percentage points**
   - **However**: Survey claims 6.66M weekly attenders vs actual church records of ~1.46M (3.7x over-reporting)
   - See [Survey Claims vs Church Records](./cofe-comparison.qmd) for detailed comparison

2. **"Never attended" decreased** from 63% (2018) to 59% (2024) - a change of **-4 percentage points**

3. **All changes are relatively small** - no category changed by more than 4 percentage points
   - **Context**: 4.2 million net migration (6.9% of E&W population) between surveys
   - Immigration could explain ~70% of the 4pp increase
   - See [Demographic Analysis](./demographic-analysis.qmd) for ONS migration data analysis

4. **Age patterns**: The youngest age group (18-34) showed the largest increase in weekly attendance (+7.0pp), compared to middle (+1.0pp) and older (+2.0pp) age groups
   - **Note**: This contradicts CofE data showing elderly (70+) make up 35% of worshippers
   - Suggests systematic over-reporting among younger respondents

5. **Internal consistency**: The 2024 binary and frequency questions show a discrepancy of `r sprintf("%.1f", abs(freq_sum_2024 - binary_2024))` percentage points

6. **Sample size**: The 2024 survey has 31% fewer respondents than 2018

7. **Demographic differences**: Ethnic minority respondents show different attendance patterns than White respondents in 2024

## Methodological Notes

**Question format**: Both surveys use the same frequency question about church attendance, making them directly comparable. The 2024 survey also includes a separate question about Bible engagement at church services, which measures a different construct.

**Sample size changes**:
- 2018: 19,875 (weighted)
- 2024: 12,455 (weighted)
- Represents a 31% reduction

**Demographic data availability**:
- Ethnicity and gender breakdowns only available for 2024
- This limits ability to assess population composition changes

## Next Steps

For critical analysis of these findings, see:

- [**Survey Claims vs Church Records**](./cofe-comparison.qmd) - **START HERE** - Comparison with actual CofE and Catholic attendance data showing 3.7x over-reporting
- [**Demographic Analysis**](./demographic-analysis.qmd) - ONS migration data showing 4.2M net migration could explain ~70% of the apparent increase
- [**Critical Analysis Overview**](./criticism-overview.qmd) - Comprehensive evaluation of methodological quality and alternative explanations
- [**Weekly Attendance Claim**](./weekly-attendance-claim.qmd) - Statistical testing of the observed changes
- [**Question Order Effects**](./question-order-effects.qmd) - Impact of methodological differences

